---
title: "COVID-19 and Community Analgesic Prescribing in Scotland"
output: html_document
date: "2025/11/27"
---

```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 9, fig.align = "center")
```

# Introduction
The COVID-19 pandemic provides a natural experiment for assessing how a major health-system disruption affected community analgesic prescribing in Scotland. Scotland’s first confirmed COVID-19 case occurred on 1 March 2020, followed by a national lockdown from 24 March 2020, which substantially changed access to and delivery of primary care, including rapid shifts to remote consultations and postponement of elective care (Scottish COVID-19 Inquiry, 2025). Legal restrictions in Scotland were largely removed by 18 April 2022, marking a transition into a “living with COVID” phase (Scottish COVID-19 Inquiry, 2025). In parallel, COVID-19 and its indirect consequences may have influenced population pain burden through delayed rehabilitation, altered injury patterns during restrictions, and post-viral symptoms such as myalgia or headache (NHS, 2025). 
This analysis uses Public Health Scotland “Prescriptions in the Community” data to describe how analgesic prescribing evolved across three non-overlapping 26-month windows, defined for comparability as:

-	Before COVID: January 2018 – February 2020  
-	During COVID: March 2020 – April 2022  
-	After COVID: May 2022 – June 2024

The primary outcome is the number of paid items for analgesics (BNF section 04.07), providing a consistent indicator of prescribing volume over time (Open Prescribing, 2025; Public Health Scotland, 2025). The analysis addresses three questions:

1. Did the overall analgesic prescribing volume change across the three periods?

2. Which specific analgesic products dominated prescribing before, during and after COVID-19, and how did their shares change?

3. Were any observed changes brief responses around the onset of the pandemic, or did they develop into longer-term shifts in practice?
To answer these questions, the report summarises the top-10 analgesic products by volume in each period, examines changes in their percentage share using a heatmap, and then evaluates monthly time-series for both total analgesic items and the union of top-10 drugs.

# Data Loading
```{r}
## Purpose: Load core packagess for data wrangling, tables, paths, and plotting
#Load libraries
library(tidyverse)
library(janitor) # cleaning data
library(gt) # nicely formatted tables
library(here) # directory structure 
library(scales) # number/percent formatting
library(glue) # string interpolation for labels
```

```{r}
## Purpose: Import data from before, during, and after COVID-19, keeping only BNF code, BNF item drscription, and number of paid items
read_cols <- function(path){read_csv(path) %>%
    clean_names() %>%
    select(any_of(c("bnf_item_code", "bnf_item_description", "number_of_paid_items")))}
# Before COVID data
FileBefore <- list.files(here("data","before"), pattern="csv", full.names=TRUE)
DataBefore <- map_dfr(FileBefore, read_cols)
# During COVID data
FileDuring <- list.files(here("data","during"), pattern="csv", full.names=TRUE)
DataDuring <- map_dfr(FileDuring, read_cols)
# After COVID data
FileAfter <- list.files(here("data","after"), pattern="csv", full.names=TRUE)
DataAfter <- map_dfr(FileAfter, read_cols)
# readr reports parsing warnings because some unused columns contain mixed or irregular values. Selected columns were checked and are clean,the warnings do not affect analysis
problems(DataBefore) %>%
  filter(col %in% c("bnf_item_code","bnf_item_description","number_of_paid_items"))
problems(DataDuring) %>%
  filter(col %in% c("bnf_item_code","bnf_item_description","number_of_paid_items"))
problems(DataAfter) %>%
  filter(col %in% c("bnf_item_code","bnf_item_description","number_of_paid_items"))
```

```{r}
## Purpose: Collapse monthly rows into period-level totals for each analgesic code (BNF04.07), and compute total analgesic volume per period
# Create a function to select BNF_item_code that begins with "0407" for before, during, and after COVID, representing all the analgesics and add in item description for corresponding BNF_item_code
AnalgesicsPeriod <- function(df, period_label) {df %>% 
    filter(startsWith(bnf_item_code, "0407")) %>%
    mutate(bnf_item_description = str_squish(bnf_item_description)) %>% 
    group_by(bnf_item_code) %>%
    summarise(paid_items = sum(number_of_paid_items, na.rm = TRUE), bnf_item_description = first(bnf_item_description), .groups = "drop") %>% 
    mutate(period = period_label)} 
# Apply to each period and stack into a table
AnalgesicsAllPeriods <- bind_rows(AnalgesicsPeriod(DataBefore, "Before COVID"), AnalgesicsPeriod(DataDuring, "During COVID"), AnalgesicsPeriod(DataAfter, "After COVID")) %>% 
  mutate(period = factor(period, levels=c("Before COVID", "During COVID", "After COVID")))
```

# Results
## Table 1.Top 10 Most Prescribed Analgesics Before, During, and After COVID-19
Table 1 shows that community analgesic prescribing in Scotland is consistently dominated by a similar set of drugs across all three periods. Paracetamol 500mg and co-codamol 30/500mg tablets remain the top two items before, during, and after COVID-19, together accounting for approximately 40-43% of all analgesic paid items in each window, indicating that first-line simple and combination analgesics remained the backbone of pain management despite major disruption to primary-care services. Tramadol 50mg capsules and dihydrocodeine 30mg tablets also remain consistently high-ranking, indicating a stable core formulary. Total analgesic volume changes little over time (18.83 million items before COVID, 18.54 million during, 19.08 million after), implying no major pandemic-related rise or fall in overall use. Instead, there’re modest compositional shifts: within the top 10, paracetamol 500mg and co-codamol 30/500mg gain about 1-1.3% in share between pre- and post-COVID periods, while tramadol 50mg capsules and co-codamol 8/500mg tablets gradually lose share. This pattern is consistent with a gradual shift away from stronger or higher-dose opioid combinations toward first-line paracetamol-based regimens, aligning with ongoing opioid-stewardship efforts in UK primary care (Levy, Lord, and Lobo, 2021). Paediatric paracetamol suspensions (120mg/5mL and 250mg/5mL) each contribute around 2-2.6% of items, with a dip during COVID and partial recovery afterwards, consistent with reduced childhood infections and GP consultations during lockdown, followed by a rebound in routine paediatric prescribing once contact rates recovered. Sumatriptan 50mg tablet maintains a low but gradually rising share, while morphine oral solution 10mg/5mL appears in the top 10 only after COVID with a small share, suggesting some intensification of community-managed severe pain in a minority of patients.

Overall, Table 1 suggests that the pandemic caused a gradual shift in the mix of drugs towards paracetamol-based regimens and away from some higher-strength opioid options, rather than a large change in total analgesic burden.  

```{r}
## Purpose: Identify the top-10 mostly prescribed analgesics in each period, compute their shares of all analgesic items, and display them in a formatted table
# Calculate total paid item quantity of analgesics during each period
Period_totals_data <- AnalgesicsAllPeriods %>%
  group_by(period) %>% 
  summarise(total_paid_items = sum(paid_items, na.rm = TRUE), .groups = "drop") 
# Top10 table
Top10_each_period_data <- AnalgesicsAllPeriods %>%
  group_by(period) %>% 
  arrange(desc(paid_items), .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  mutate(rank = row_number()) %>%
  ungroup() %>% 
  left_join(Period_totals_data, by = "period") %>% 
  mutate(share_pct = paid_items/total_paid_items, period = factor(period, levels = c("Before COVID","During COVID","After COVID")), period_label = glue("{period} (Total analgesics: {comma(total_paid_items)})"))
  # Make this table pretty 
Top10_each_period_table <- Top10_each_period_data %>% 
  select(period, period_label, rank, bnf_item_code, bnf_item_description, paid_items, share_pct) %>% 
  arrange(period, rank) %>% 
  gt(groupname_col = "period_label") %>%
  tab_header(title = md("Top 10 Most Prescribed Analgesics Before, During, and After COVID-19")) %>%
  cols_hide(columns = period) %>% 
  cols_label(rank = "Rank", bnf_item_code = "BNF Item Code", bnf_item_description = "BNF Item Description", paid_items = "Number of Paid Items", share_pct = "Share of Analgesics(%)") %>%
  fmt_number(columns = paid_items, decimals = 0, use_seps = TRUE) %>%
  fmt_percent(columns = share_pct, decimals = 2) %>% 
  cols_width(rank ~ px(60), bnf_item_code ~ px(140), bnf_item_description ~ px(340), paid_items ~ px(140), share_pct ~ px(140)) %>%
  cols_align(align = "center", columns = c(rank, bnf_item_code, paid_items, share_pct)) %>%
  cols_align(align = "left", columns = bnf_item_description) %>%
  opt_row_striping() %>%
  tab_style(style = list(cell_fill(color = "#F5F5F5"), cell_text(weight = "bold")), locations = cells_body(columns = rank)) %>%
  tab_options(row_group.border.top.style = "solid", row_group.border.top.width = px(2), row_group.border.top.color = "black", row_group.padding = px(6)) %>%
  tab_options(table.font.size = px(13), heading.title.font.size = px(16), heading.subtitle.font.size = px(11), row_group.font.weight = "bold", row_group.font.size = px(12), column_labels.font.weight = "bold", column_labels.border.bottom.width = px(2), table.border.top.width = px(2), table.border.bottom.width = px(2), data_row.padding = px(4))
Top10_each_period_table
```

## Figure 1. Change in Share of Union Top10 Analgesics Prescription
To examine these compositional changes more directly, Figure 1 presents a heatmap of percentage-point changes in share for the union of Top-10 drugs (“During vs Before”, “After vs During”, “After vs Before”). Warm cells indicate gains in share and cool cells indicate losses. 
The pattern in Figure 1 mirrors and extends Table 1. Paracetamol 500mg and co-codamol 30/500mg tablets show the largest early increases in share between the pre-COVID and COVID windows, and these gains are sustained into the post-COVID period. In contrast, high-dose combinations that fall in Table 1 (i.e. tramadol 50mg, co-dydramol 10/500mg and co-codamol 8/500mg) display consistent losses across both transitions. The heatmap also makes visible drugs with small absolute shares but marked relative shifts, such as paediatric paracetamol suspensions, which show a clear trough during COVID and rebound afterwards. 
Taken together, Table 1 and Figure 1 indicate a gradual reweighting of prescribing toward simple and lower-dose combination analgesics and away from stronger opioid formulations, without the emergence of entirely new leading products.
```{r}
## Track how the share of each union top-10 drug changes between periods and visualise the percentage-point differences as a heatmap
# Union of Top-10 analgesic codes across all periods
top10_union <- Top10_each_period_data %>%
  distinct(bnf_item_code, bnf_item_description)
# Pull paid_items for union codes from full period data, then compute share
union_shares <- AnalgesicsAllPeriods %>%
  semi_join(top10_union, by = "bnf_item_code") %>%
  left_join(Period_totals_data, by = "period") %>%
  mutate(share = paid_items / total_paid_items) %>%
  select(period, bnf_item_code, bnf_item_description, share) %>%
  pivot_wider(names_from  = period, values_from = share, values_fill = 0)
# Contrast the percentage share of the union top-10 drugs before vs during vs after COVID
union_change <- union_shares %>%
  mutate(`During vs Before` = `During COVID` - `Before COVID`,
         `After vs During`  = `After COVID`  - `During COVID`,
         `After vs Before`  = `After COVID`  - `Before COVID`) %>%
  pivot_longer(cols = c(`During vs Before`,`After vs During`,`After vs Before`), names_to = "contrast", values_to = "delta") 
# Order rows by maximum absolute share change across contrasts to highlight drugs with the largest compositional shifts
order_levels <- union_change %>%
  group_by(bnf_item_description) %>%
  summarise(max_abs = max(abs(delta)), .groups="drop") %>%
  arrange(desc(max_abs)) %>%
  pull(bnf_item_description)
# Heatmap
union_change_plot <- union_change %>% 
  mutate(contrast = factor(contrast, levels = c("During vs Before", "After vs During", "After vs Before")),
         bnf_item_description = factor(bnf_item_description, levels = rev(order_levels)), drug_label = str_wrap(bnf_item_description, width = 28))
# Symmetric colour limits around 0
lim <- max(abs(union_change_plot$delta), na.rm = TRUE)
ggplot(union_change_plot, aes(x = contrast, y = drug_label, fill = delta)) +
  geom_tile(color = "white", linewidth = 0.6) +
  geom_text(aes(label = percent(delta, accuracy = 0.1)), size = 3.2, fontface = "bold") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b", limits = c(-lim, lim), breaks = pretty_breaks(5), labels = percent_format(accuracy = 0.1), name = "Δ share")+
  scale_x_discrete(position = "top") +
  labs(title = "Change in Share of Union Top10 Analgesics Prescription",
       subtitle = "Union of Top-10 drugs across periods (values are percentage_point changes)", x = NULL, y = NULL, caption = "Warm colours = increased share; Cool colours = decreased share.") + theme_minimal(base_size = 12) + theme(panel.grid = element_blank(), axis.text.x = element_text(face = "bold", size = 10), axis.text.y = element_text(size = 9), plot.title = element_text(face = "bold", size = 15, hjust = 0.5), plot.subtitle = element_text(size = 10, hjust = 0.5, margin = margin(b=8)), legend.title = element_text(face = "bold"), legend.key.height = unit(1.1, "cm"), plot.margin = margin(10, 12, 10, 10))
```

## Figure 2. Monthly Presciption of All Analgesics in Scotland 
To assess whether changes in drug share reflected redistribution within a stable total volume or genuine shifts in overall analgesic use, Figure 2 plots monthly totals for all BNF 04.07 items from before COVID to after COVID, with the COVID restriction period shaded. The series shows a pronounced peak around March 2020 (i.e. start of COVID-19), followed by a temporary reduction during the restriction period and a steady recovery from 2022 onwards, reaching slightly higher levels than in the immediate pre-pandemic years. 
The short-lived surge at the onset of COVID-19 may reflect anticipatory stockpiling. The mid-pandemic contraction is consistent with reduced primary-care contact and fewer non-COVID illnesses. The later rebound and slight may reflect catch-up of deferred care and ongoing pain needs in patients whose treatment or procedures were delayed.

```{r}
## Purpose: Show overall monthly volume of all analgesics across the study period
# Load Monthly Analgesics Data
read_cols_month <- function(path, period_label){
  yyyymm <- str_extract(basename(path), "\\d{6}") 
  
read_csv(path, show_col_types = FALSE) %>%
  clean_names() %>%
  select("bnf_item_code", "bnf_item_description", "number_of_paid_items") %>%
  mutate(covid_period = period_label, month  = yyyymm, bnf_item_description = str_squish(bnf_item_description)) %>% 
  filter(startsWith(bnf_item_code, "0407"))}

load_monthly <- function(folder, period_label){
  files <- list.files(here("data", folder), pattern = "csv", full.names = TRUE)
  map_dfr(files, read_cols_month, period_label = period_label)}

AllMonthly <- bind_rows(
  load_monthly("before", "Before COVID"),
  load_monthly("during", "During COVID"),
  load_monthly("after",  "After COVID")) %>%
  mutate(covid_period = factor(covid_period, levels = c("Before COVID","During COVID","After COVID")), month = as.character(month)) %>%
  filter(!is.na(bnf_item_code), !is.na(bnf_item_description), !is.na(number_of_paid_items), !is.na(month))
# To confirm parsing issue unrelated
problems(AllMonthly) %>%
  filter(col %in% c("bnf_item_code","bnf_item_description","number_of_paid_items", "month"))
# Monthly totals for analgesics Figure
Analgesics_monthly_totals <- AllMonthly %>%
  group_by(month) %>%
  summarise(total_paid_items = sum(number_of_paid_items, na.rm = TRUE), .groups = "drop") %>% mutate(month_date = ym(month))
# Define periods on graph
covid_start <- ym("2020-03")
covid_end <- ym("2022-05")
covid_band <- tibble(xmin = covid_start, xmax = covid_end, ymin = -Inf, ymax = Inf)
# Plot Monthly Presciption of All Analgesics in Scotland
ggplot(Analgesics_monthly_totals, aes(x = month_date, y = total_paid_items)) +
  geom_rect(data = covid_band, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), inherit.aes = FALSE, fill = "grey85", alpha = 0.5) +
  geom_line(linewidth = 0.8, colour = "#2C6BAA") +
  geom_smooth(method = "loess", se = FALSE, colour = "black") +
  scale_x_date(date_breaks="3 months", date_labels="%Y-%m") +
  scale_y_continuous(labels=comma) +
  labs(title = "Monthly Presciption of All Analgesics in Scotland (BNF 04.07)",
       subtitle = "Grey band = COVID restriction period; black curve = LOESS trend",
    x = "Year-Month", y = "Total Paid Items") +
  theme_minimal(base_size = 14) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title.position = "plot",
        plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
        plot.subtitle = element_text(size = 10, margin = margin(b = 6), hjust = 0.5))
```

## Figure 3. Monthly Prescription of Union Top10 Analgesics
To link these aggregate patterns to individual drugs, Figure 3 plots monthly counts for the union of Top-10 analgesics, each with its own LOESS trend (black curve) summarising the underlying trend by averaging over month-to-month noise, and shaded COVID period. 
Paracetamol 500mg and co-codamol 30/500mg tablets show clear spikes at the start of COVID and then settle onto higher post-pandemic baselines, matching their increased post-COVID shares in Figure 1. Tramadol 50mg capsules, co-dydramol 10/500 mg and co-codamol 8/500 mg tablets display steady downward trajectories with only minor perturbation during COVID, consistent with their declining shares on the heatmap and the small falls in Table 1. 
Paediatric paracetamol suspensions (120mg/5mL and 250mg/5mL) exhibit pronounced month-to-month variability but share a visible trough within the COVID band followed by recovery or overshoot, supporting the interpretation of reduced paediatric consultations and infections during restrictions and rebound afterwards. Codeine 30mg, dihydrocodeine 30mg, and sumatriptan 50mg tablets and morphine oral solution 10mg/5mL all show gradual upward trends, with morphine mainly rising in the later, post-COVID phase. Together with Figure 2, these facets suggest a brief disturbance in overall analgesic volume during COVID, superimposed on longer-term shifts away from some higher-risk opioid combinations and towards paracetamol-based and selected alternative agents.
```{r}
## Purpose: Examine drug-specific trajectories for the union of top-10 analgesics
# Union of Top10 analgesic codes across all periods, and a label for plotting
code_labels <- union_shares %>% 
  distinct(bnf_item_code, bnf_item_description) %>% 
  mutate(bnf_item_description = str_squish(bnf_item_description), label = bnf_item_description)
top10_union_codes <- code_labels %>% 
  pull(bnf_item_code)
# Monthly totals for Union drugs only
top10_monthly_code <- AllMonthly %>%
 filter(bnf_item_code %in% top10_union_codes) %>%
  mutate(bnf_item_description = str_squish(bnf_item_description), month_date = ym(month)) %>%
  group_by(month_date, bnf_item_code) %>%
  summarise(paid_items = sum(number_of_paid_items, na.rm = TRUE), .groups = "drop") %>%
  left_join(code_labels, by = "bnf_item_code")
# Plot Monthly Prescription of Union Top10 Analgesics
ggplot(top10_monthly_code, aes(x = month_date, y = paid_items)) +
  geom_rect(data = covid_band, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), inherit.aes = FALSE, fill = "grey85", alpha = 0.5) +
  geom_line(linewidth = 0.7, colour = "#2C6BAA") +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.6, colour = "black")+
  facet_wrap(~ label, scales = "free_y", ncol = 2, labeller = label_wrap_gen(width = 22)) +
  scale_x_date(date_breaks = "6 months", date_labels = "%Y-%m") +
  scale_y_continuous(labels = comma, breaks = scales::pretty_breaks(4)) +
  labs(title = "Monthly Prescription for Union of Top-10 Analgesics", subtitle = "Grey band = COVID restriction period; black curve = within-drug LOESS trend", x = "Year-Month", y = "Number of paid items")+
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), panel.grid.major.x = element_blank(), strip.text = element_text(face = "bold", size = 10, margin = margin(b=2)), strip.background = element_rect(fill = NA, colour = NA), panel.spacing = unit(0.9, "lines"), axis.text.x = element_text(size = 8, angle = 45, hjust = 1), plot.title.position = "plot", plot.title = element_text(face = "bold", size = 15, hjust = 0.5), plot.subtitle = element_text(size = 10, margin = margin(b = 6), hjust = 0.5))
```

# Conclusion
This analysis suggests that COVID-era changes in Scottish community analgesic prescribing were modest in volume but meaningful in composition. Total paid items for analgesics were very similar across the three 26-month windows, with only a brief spike around March 2020, a mid-pandemic dip, and a modest recovery above pre-COVID levels. In contrast, the mix of drugs changed: prescribing remained concentrated in a small set of agents, but within that group paracetamol 500mg and co-codamol 30/500mg became slightly more prominent, while tramadol 50mg and other higher-dose opioid combinations declined. Paediatric paracetamol suspensions showed a trough during restrictions and subsequent rebound, consistent with temporary reductions in childhood infections and primary-care contact. Together, the time-series and share analyses indicate that early-pandemic disturbances were short-lived, whereas the gradual move away from some opioid formulations and toward paracetamol-based regimens represents a longer-term shift that continued beyond the main restriction period.
These findings are consistent with ongoing opioid-stewardship efforts in Scottish primary care, with no evidence of a sustained COVID-driven surge in strong-opioid use. However, several limitations qualify the interpretation. 
Paid items capture dispensing events but not dose, treatment duration, indication, or over-the-counter and hospital use. The analysis focuses on aggregate data and the union of the top-10 drugs, so it cannot reveal patient-level differences by age, deprivation or diagnosis, nor trends in less commonly prescribed agents. Collapsing a complex pandemic into three broad periods also obscures shorter policy phases and regional variation, and the descriptive design cannot establish causality.
Future work could extend the time horizon to see whether post-COVID patterns persist or attenuate, and link community prescribing to richer datasets such as demographic information, chronic pain registers, procedure and discharge data, or infection surveillance to examine how trends differ across clinical and social groups. Using dose-based metrics (e.g. defined daily doses or oral morphine equivalents) and applying interrupted time-series methods around key policy and pandemic milestones would allow more formal assessment of changes in true analgesic exposure and stewardship impact. The workflow used here could also be applied to other drug classes to monitor how future system shocks or policy interventions shape community prescribing.

# References
Levy, N., Lord, L.J. and Lobo, D.N. (2021). ‘UK recommendations on opioid stewardship’, BMJ, 372, pp.m4901. Available at:https://doi.org/10.1136/bmj.m4901.

NHS (2025) Long COVID. Available at:https://www.nhs.uk/conditions/long-covid/#:~:text=a%20high%20temperature,anxiety%20and%20depression (Accessed: 26 November 2025).

Open Prescribing (2025) All BNF sections. Available at:https://openprescribing.net/bnf/ (Accessed: 26 November 2025).

Public Health Scotland (2025) Prescriptions in the Community. Available at:https://www.opendata.nhs.scot/dataset/prescriptions-in-the-community (Accessed: 26 November 2025). 

Scottish COVID-19 Inquiry (2025) COVID-19 Pandemic in Scotland - a timeline of key dates. Available at:https://www.covid19inquiry.scot/covid-19-pandemic-scotland-timeline-key-dates (Accessed: 26 November 2025). 


